\huge{Meta-analytic evidence that sexual selection improves population fitness}

### 2019-03-28 Manuscripts/Manuscript_MetaAnalysis.Rmd
Manuscript updates

We retrieved 459 effect sizes from 65 studies. 92 effect sizes were collected from populations evolving under stressful conditions, while 337 were measured for those evolving in benign conditions. 189 of the effect sizes came from measurements made on males, 219 on females, and the remaining 51 from measurements of a mixed-sex sample of individuals. Most effect sizes in our dataset came from studies that manipulated sexual selection by completely removing it in one treatment via enforced random monogamy ($n$ = 241); other effect sizes ($n$ = 218) derived from alternative manipulations, such as changing the adult sex ratio. In total, we obtained effect sizes for 22 different fitness traits, with female reproductive success ($n$ = 102) and offspring viability ($n$ = 56) being the most commonly-measured traits. We classified 171 effect sizes as direct measures of population fitness, 141 as indirect, and the remaining 144 effect sizes as ambiguous (see Methods, \autoref{forest_plot} and \href{https://justincally.github.io/SexualSelection/#inclusionexclusion_criteria}{Supplementary Table 1}). Specifically, we scored traits that are likely to correlate with population growth and persistence as direct (*e.g.*, female reproductive success, offspring viability, and extinction rate), those which are not necessarily correlated with population fitness but which do measure individual fitness as indirect (*e.g.*, lifespan, male mating success and ejaculate quality/production); and those for which the relationship to population fitness is unclear as ambiguous (*e.g.*, body size, mating duration, male reproductive success, and early fecundity). \href{https://justincally.github.io/SexualSelection/#tables_of_sample_sizes}{Supplementary Tables 3,4} give a detailed description of our dataset.\phantomsection

@@

The grand mean across all types of effect sizes (direct, indirect and ambiguous) was positive (REML $\beta$ = 0.24, 95% CIs: 0.055 to 0.43, $p$ = 0.011; Bayesian $\beta$ = 0.25, 95% CIs: -0.0074 to 0.51, $BF_{>0}$ = 35), indicating that sexual selection on males typically had a net positive effect on the majority of populations and fitness components so far studied. Moreover, the effect sizes associated with the manipulation of sexual selection varied between different fitness traits. Sexual selection had a beneficial effect on most fitness traits, but varied across the three relationships to fitness \autoref{forest_plot}; \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_on_direct,_indirect_and_ambiguous_measures_of_fitness}{Supplementary Table 6}). Sexual selection elevated fitness for traits that shared an ambiguous relationship to fitness (REML $\beta$ = 0.21, 95% CIs: 0.058 to 0.093; Bayesian $\beta$ = 0.20, 95% CIs: -0.0016 to	0.39, $BF_{>0}$ = 38, $n$ = 144) and an indirect relationship to fitness (REML $\beta$ = 0.24, 95% CIs: 0.13 to 0.36; Bayesian $\beta$ = 0.24, 95% CIs: 0.033 to 0.43, $BF_{>0}$ = 59, $n$ = 141). Additionally, sexual selection elevated fitness components directly related to fitness, albeit at a lower magnitude (REML $\beta$ = 0.13, 95% CIs: 0.019 to 0.24; Bayesian $\beta$ = 0.13, 95% CIs: -0.079 to 0.31, $BF_{>0}$ = 11, $n$ = 174). A large forest plot with predicted effect sizes for each fitness component is presented in \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Supplementary Figure 1} and further detailed with model predictions and individual meta-analyses in \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Supplementary Table 8,9}. Sexual selection significantly reduced two fitness components, namely immunity (REML $\beta$ = -0.42, 95% CIs: -0.64 to -0.20; Bayesian $\beta$ = -0.43, 95% CIs: -0.70 to -0.15, $BF_{>0}$ = 0.0026; $n$ = 35) and body condition (REML $\beta$ = -1.2, 95% CIs: -1.8 to -0.63; Bayesian $\beta$ = -1.2, 95% CIs: -1.9 to -0.63, $BF_{>0}$ < 0.0001, $n$ = 1).  \phantomsection

@@

We found that the sex of the individuals measured (male, female, or a mixture), and the conditions under which the population evolved (stressful or benign) interacted to affect the relationship between sexual selection and fitness (Table 1,2). Sexual selection on males significantly improved female fitness, and the beneficial effect of sexual selection was significantly stronger for females from populations evolving under stressful conditions (*e.g.* a food source to which they were not well-adapted) than under benign conditions (\autoref{env_sex}a, Table 2). Sexual selection had a positive but non-significant effect on male fitness, and in contrast to females, fitness benefits were significantly weaker in stressful than benign environments (\autoref{env_sex}a, Table 2). Consistent with the different consequences of sexual selection for female and male fitness, the mean effect size in mixed-sex samples was non-significantly positive, and there was no significant difference between benign versus stressful conditions (\autoref{env_sex}a, Table 2). Overall, our results indicate that the positive effect of sexual selection on fitness is greater for females than males, and the difference between the sexes is magnified in stressful environments. When we used an alternative measure of effect size (log response ratio or lnRR), the results aligned with those of Hedges' g in that sexual selection elevates population fitness, with its effect magnified for females evolving in stressful environments (\href{https://justincally.github.io/SexualSelection/#meta-analysis_using_lnrr}{Supplementary Figure 4 and Supplementary Table 16}).

@@

By applying meta-analysis to log coefficient of variation ratios [*lnCVR*; @RN177], we found evidence that sexual selection reduces phenotypic variation under certain conditions (\autoref{env_sex}b). Specifically, phenotypic variance was significantly reduced by sexual selection for fitness components measured in females under stressful conditions (*lnCVR* = -0.78, 95% CIs: -1.23 to -0.34, $n$ = 27). By contrast, we found no significant effect of sexual selection on phenotypic variance in males, or for either sex under benign conditions (\autoref{env_sex}b; \href{https://justincally.github.io/SexualSelection/#meta-analysis_on_variance}{Supplementary Tables 17-20}). However, similar to the results in females, there was a non-significant trend for a reduction in phenotypic variance in mixed-sex samples measured under stressful conditions (*lnCVR* = -0.76, 95% CIs: -1.22 to -0.31; \autoref{env_sex}b). A meta-analysis using the log variability ratio (lnVR), which does not account for the mean-variance relationship present in the dataset (\href{https://justincally.github.io/SexualSelection/#lnvr_effect_size_of_variance}{Supplementary Figure 5}), suggested sexual selection reduces variance for mixed-sex samples in stressful conditions, but not females, results of this meta-analysis can be found in the Supplementary Information (\href{https://justincally.github.io/SexualSelection/#lnvr_effect_size_of_variance}{Supplementary Figure 6 and Supplementary Table 21}).

@@

The funnel plot of effect sizes was asymmetrical, suggesting that some publication bias might be present (\autoref{publication_bias}a; Egger’s test: *z* = 5.9, $p$ < 0.0001). Specifically, there was a moderate excess of low-powered studies in which sexual selection had a more positive effect on the fitness component than average, implying that low-powered studies are more likely to be published if they report statistically significant fitness benefits of sexual selection (though funnel plots are not a decisive evidence of publication bias[@RN215]). Linear regressions show no significant relationship between effect size and journal impact factor (\autoref{publication_bias}b; $t_{437}$ = 1.2, $p$ = 0.23) or year of publication (\autoref{publication_bias}c; $t_{437}$ = -1.2, $p$ = 0.24); thus, we found no evidence that effect size dictates the likelihood of publication in high-profile journals, or that effect sizes have diminished as the field has matured [@RN216; @RN217].

@@

After removing duplicates, we read the titles and abstracts of the remaining 1,015 papers and removed those that did not fit our inclusion criteria (typically because they did not present primary experimental evolution data). This left 130 papers, for which we read the full text and applied the inclusion criteria outlined in the PRISMA diagram (\autoref{prisma_flow}). Briefly, we included studies that 1) were conducted in a dioecious animal, 2) experimentally manipulated the strength of sexual selection (*e.g.*, via experimentally-enforced random monogamy or an altered sex ratio) for at least one generation, and 3) measured a trait that we judged to be a potential correlate of population fitness. This third criterion is the most subjective, because there is rarely enough data to determine whether a particular trait is (or is not) correlated with population fitness. We therefore relied on our best judgement when deciding what outcomes were correlated with population fitness. We categorised the fitness outcomes into three categories: ambiguous, indirect and direct (detailed in \href{https://justincally.github.io/SexualSelection/#inclusionexclusion_criteria}{Supplementary Table 1}). Briefly, ambiguous measures of fitness were those that are reported to have an unclear or variable association with fitness (*e.g.*, body size, mating duration, early fecundity and male reproductive success). Indirect fitness components were those that are often used as a proxy of fitness but do not directly measure aspects of success in reproduction or population viability (*e.g.*, lifespan, mating success and ejaculate quality/production). Finally, direct measures of fitness (female/mixed sex reproductive success, offspring viability and extinction rate) are those that measure fitness through components of reproduction or long-term viability. The Supplementary Methods describe why each of the 130 papers was included or excluded (\href{https://justincally.github.io/SexualSelection/#inclusionexclusion_criteria}{Supplementary Table 2}).

Of these 130 papers, 62 were excluded based on the PRISMA criteria (\autoref{prisma_flow}). Additionally, three papers presented insufficient information to calculate effect size. In these cases, we contacted the authors and attempted to obtain the missing data, with partial success. The final meta-analysis included data from 65 papers.

@@

![\textbf{The effect of sexual selection on direct, indirect and ambiguous fitness components.} The effect sizes used in this meta-analysis ($n$ = 459) were grouped into either direct, indirect or ambiguous measures of fitness. Overall, effect sizes were more often positive than negative. Predicted average values are presented as a diamond with 95 % CIs for each fitness-relationship category. The estimates presented here are from REML models with the grand mean across all effect sizes ($\beta$ = 0.25) shown as the blue dotted line. Predictions from both Bayesian and REML models can be found in Supplementary Table 6 and source data are provided at \url{https://github.com/JustinCally/SexualSelection}.\label{forest_plot}](Figures/forest.plot.class.pdf)

![\textbf{The roles of environmental stress and sex.} (a) Sexual selection tends to increase the population mean values of fitness traits, especially for female traits and for populations living under stressful conditions. (b) Under stressful conditions, sexual selection tends to reduce the phenotypic variance in fitness traits, especially for traits measured in females or mixed-sex individuals. The points with error bars show the mean effect sizes and their 95% CIs, determined from a meta-regression fit using restricted maximum likelihood (REML); the point sizes are proportional to the number of effect sizes (see Supplementary Tables 14,18). Results from Bayesian meta-regression are shown as posterior prediction density curves, with vertical lines indicating the median. Source data are provided at \url{https://github.com/JustinCally/SexualSelection}.\label{env_sex}](Figures/environment_variance_edit5.pdf)

![\textbf{Tests for publication bias in the dataset.} Tests of publication bias are mixed and suggest publication bias may be present. (a) Inspection and statistical tests of the funnel plot reveal large amounts of heterogeneity in the dataset with asymmetry from increased low-powered, large effect studies. (b) No significant correlation exists between journal impact factor and effect size, (c) additionally there is no significant correlation between effect size and year of publication when testing for the time-lag bias. For (b) and (c) point size is proportional to the precision of the effect size (*i.e.* the inverse of its variance). For (a), the dashed red lines represent the 95 % CIs of the grand mean estimate for all effect sizes and the black dotted and dashed lines depict the 95 % and 99.8 % CIs for the dataset. The grey envelopes in (b) and (c) represent the 95 % CI of the linear regression. Source data are provided at \url{https://github.com/JustinCally/SexualSelection}.\label{publication_bias}](Figures/publicationbias.pdf)

![\textbf{PRISMA diagram.} Flow of inclusion and exclusion of studies identified during the literature search, presented as a PRISMA diagram with number of published papers in brackets.\label{prisma_flow}](Figures/PRISMA\ Chart.pdf)

>>>>

We retrieved 459 effect sizes from 65 studies. 92 effect sizes were collected from populations evolving under stressful conditions, while 337 were measured for those evolving in benign conditions. 189 of the effect sizes came from measurements made on males, 219 on females, and the remaining 51 from measurements of a mixed-sex sample of individuals. Most effect sizes in our dataset came from studies that manipulated sexual selection by completely removing it in one treatment via enforced random monogamy ($n$ = 241); other effect sizes ($n$ = 218) derived from alternative manipulations, such as changing the adult sex ratio. In total, we obtained effect sizes for 22 different fitness traits, with female reproductive success ($n$ = 102) and offspring viability ($n$ = 56) being the most commonly-measured traits. We classified 171 effect sizes as direct measures of population fitness, 141 as indirect, and the remaining 144 effect sizes as ambiguous (see Methods, Figure 1 and \href{https://justincally.github.io/SexualSelection/#inclusionexclusion_criteria}{Supplementary Table 1}). Specifically, we scored traits that are likely to correlate with population growth and persistence as direct (*e.g.*, female reproductive success, offspring viability, and extinction rate), those which are not necessarily correlated with population fitness but which do measure individual fitness as indirect (*e.g.*, lifespan, male mating success and ejaculate quality/production); and those for which the relationship to population fitness is unclear as ambiguous (*e.g.*, body size, mating duration, male reproductive success, and early fecundity). \href{https://justincally.github.io/SexualSelection/#tables_of_sample_sizes}{Supplementary Tables 3,4} give a detailed description of our dataset.\phantomsection

@@

The grand mean across all types of effect sizes (direct, indirect and ambiguous) was positive (REML $\beta$ = 0.24, 95% CIs: 0.055 to 0.43, $p$ = 0.011; Bayesian $\beta$ = 0.25, 95% CIs: -0.0074 to 0.51, $BF_{>0}$ = 35), indicating that sexual selection on males typically had a net positive effect on the majority of populations and fitness components so far studied. Moreover, the effect sizes associated with the manipulation of sexual selection varied between different fitness traits. Sexual selection had a beneficial effect on most fitness traits, but varied across the three relationships to fitness (Figure 1; \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_on_direct,_indirect_and_ambiguous_measures_of_fitness}{Supplementary Table 6}). Sexual selection elevated fitness for traits that shared an ambiguous relationship to fitness (REML $\beta$ = 0.21, 95% CIs: 0.058 to 0.093; Bayesian $\beta$ = 0.20, 95% CIs: -0.0016 to	0.39, $BF_{>0}$ = 38, $n$ = 144) and an indirect relationship to fitness (REML $\beta$ = 0.24, 95% CIs: 0.13 to 0.36; Bayesian $\beta$ = 0.24, 95% CIs: 0.033 to 0.43, $BF_{>0}$ = 59, $n$ = 141). Additionally, sexual selection elevated fitness components directly related to fitness, albeit at a lower magnitude (REML $\beta$ = 0.13, 95% CIs: 0.019 to 0.24; Bayesian $\beta$ = 0.13, 95% CIs: -0.079 to 0.31, $BF_{>0}$ = 11, $n$ = 174). A large forest plot with predicted effect sizes for each fitness component is presented in \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Supplementary Figure 1} and further detailed with model predictions and individual meta-analyses in \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Supplementary Table 8,9}. Sexual selection significantly reduced two fitness components, namely immunity (REML $\beta$ = -0.42, 95% CIs: -0.64 to -0.20; Bayesian $\beta$ = -0.43, 95% CIs: -0.70 to -0.15, $BF_{>0}$ = 0.0026; $n$ = 35) and body condition (REML $\beta$ = -1.2, 95% CIs: -1.8 to -0.63; Bayesian $\beta$ = -1.2, 95% CIs: -1.9 to -0.63, $BF_{>0}$ < 0.0001, $n$ = 1).  \phantomsection

@@

We found that the sex of the individuals measured (male, female, or a mixture), and the conditions under which the population evolved (stressful or benign) interacted to affect the relationship between sexual selection and fitness (Table 1,2). Sexual selection on males significantly improved female fitness, and the beneficial effect of sexual selection was significantly stronger for females from populations evolving under stressful conditions (*e.g.* a food source to which they were not well-adapted) than under benign conditions (Figure 2a, Table 2). Sexual selection had a positive but non-significant effect on male fitness, and in contrast to females, fitness benefits were significantly weaker in stressful than benign environments (Figure 2a, Table 2). Consistent with the different consequences of sexual selection for female and male fitness, the mean effect size in mixed-sex samples was non-significantly positive, and there was no significant difference between benign versus stressful conditions (Figure 2a, Table 2). Overall, our results indicate that the positive effect of sexual selection on fitness is greater for females than males, and the difference between the sexes is magnified in stressful environments. When we used an alternative measure of effect size (log response ratio or lnRR), the results aligned with those of Hedges' g in that sexual selection elevates population fitness, with its effect magnified for females evolving in stressful environments (\href{https://justincally.github.io/SexualSelection/#meta-analysis_using_lnrr}{Supplementary Figure 4 and Supplementary Table 16}).

@@

By applying meta-analysis to log coefficient of variation ratios [*lnCVR*; @RN177], we found evidence that sexual selection reduces phenotypic variation under certain conditions (Figure 2b). Specifically, phenotypic variance was significantly reduced by sexual selection for fitness components measured in females under stressful conditions (*lnCVR* = -0.78, 95% CIs: -1.23 to -0.34, $n$ = 27). By contrast, we found no significant effect of sexual selection on phenotypic variance in males, or for either sex under benign conditions (Figure 2b; \href{https://justincally.github.io/SexualSelection/#meta-analysis_on_variance}{Supplementary Tables 17-20}). However, similar to the results in females, there was a non-significant trend for a reduction in phenotypic variance in mixed-sex samples measured under stressful conditions (*lnCVR* = -0.76, 95% CIs: -1.22 to -0.31; Figure 2b). A meta-analysis using the log variability ratio (lnVR), which does not account for the mean-variance relationship present in the dataset (\href{https://justincally.github.io/SexualSelection/#lnvr_effect_size_of_variance}{Supplementary Figure 5}), suggested sexual selection reduces variance for mixed-sex samples in stressful conditions, but not females, results of this meta-analysis can be found in the Supplementary Information (\href{https://justincally.github.io/SexualSelection/#lnvr_effect_size_of_variance}{Supplementary Figure 6 and Supplementary Table 21}).

@@

The funnel plot of effect sizes was asymmetrical, suggesting that some publication bias might be present (Figure 3a; Egger’s test: *z* = 5.9, $p$ < 0.0001). Specifically, there was a moderate excess of low-powered studies in which sexual selection had a more positive effect on the fitness component than average, implying that low-powered studies are more likely to be published if they report statistically significant fitness benefits of sexual selection (though funnel plots are not a decisive evidence of publication bias[@RN215]). Linear regressions show no significant relationship between effect size and journal impact factor (Figure 3b; $t_{437}$ = 1.2, $p$ = 0.23) or year of publication (Figure 3c; $t_{437}$ = -1.2, $p$ = 0.24); thus, we found no evidence that effect size dictates the likelihood of publication in high-profile journals, or that effect sizes have diminished as the field has matured [@RN216; @RN217].

@@

After removing duplicates, we read the titles and abstracts of the remaining 1,015 papers and removed those that did not fit our inclusion criteria (typically because they did not present primary experimental evolution data). This left 130 papers, for which we read the full text and applied the inclusion criteria outlined in the PRISMA diagram (Figure 4). Briefly, we included studies that 1) were conducted in a dioecious animal, 2) experimentally manipulated the strength of sexual selection (*e.g.*, via experimentally-enforced random monogamy or an altered sex ratio) for at least one generation, and 3) measured a trait that we judged to be a potential correlate of population fitness. This third criterion is the most subjective, because there is rarely enough data to determine whether a particular trait is (or is not) correlated with population fitness. We therefore relied on our best judgement when deciding what outcomes were correlated with population fitness. We categorised the fitness outcomes into three categories: ambiguous, indirect and direct (detailed in \href{https://justincally.github.io/SexualSelection/#inclusionexclusion_criteria}{Supplementary Table 1}). Briefly, ambiguous measures of fitness were those that are reported to have an unclear or variable association with fitness (*e.g.*, body size, mating duration, early fecundity and male reproductive success). Indirect fitness components were those that are often used as a proxy of fitness but do not directly measure aspects of success in reproduction or population viability (*e.g.*, lifespan, mating success and ejaculate quality/production). Finally, direct measures of fitness (female/mixed sex reproductive success, offspring viability and extinction rate) are those that measure fitness through components of reproduction or long-term viability. The Supplementary Methods describe why each of the 130 papers was included or excluded (\href{https://justincally.github.io/SexualSelection/#inclusionexclusion_criteria}{Supplementary Table 2}).

Of these 130 papers, 62 were excluded based on the PRISMA criteria (Figure 4). Additionally, three papers presented insufficient information to calculate effect size. In these cases, we contacted the authors and attempted to obtain the missing data, with partial success. The final meta-analysis included data from 65 papers.

@@

\textbf{Figure 1: The effect of sexual selection on direct, indirect and ambiguous fitness components.} The effect sizes used in this meta-analysis ($n$ = 459) were grouped into either direct, indirect or ambiguous measures of fitness. Overall, effect sizes were more often positive than negative. Predicted average values are presented as a diamond with 95 % CIs for each fitness-relationship category. The estimates presented here are from REML models with the grand mean across all effect sizes ($\beta$ = 0.25) shown as the blue dotted line. Predictions from both Bayesian and REML models can be found in Supplementary Table 6 and source data are provided at \url{https://github.com/JustinCally/SexualSelection}.

\textbf{Figure 2: The roles of environmental stress and sex.} (a) Sexual selection tends to increase the population mean values of fitness traits, especially for female traits and for populations living under stressful conditions. (b) Under stressful conditions, sexual selection tends to reduce the phenotypic variance in fitness traits, especially for traits measured in females or mixed-sex individuals. The points with error bars show the mean effect sizes and their 95% CIs, determined from a meta-regression fit using restricted maximum likelihood (REML); the point sizes are proportional to the number of effect sizes (see Supplementary Tables 14,18). Results from Bayesian meta-regression are shown as posterior prediction density curves, with vertical lines indicating the median. Source data are provided at \url{https://github.com/JustinCally/SexualSelection}.

\textbf{Figure 3: Tests for publication bias in the dataset.} Tests of publication bias are mixed and suggest publication bias may be present. (a) Inspection and statistical tests of the funnel plot reveal large amounts of heterogeneity in the dataset with asymmetry from increased low-powered, large effect studies. (b) No significant correlation exists between journal impact factor and effect size, (c) additionally there is no significant correlation between effect size and year of publication when testing for the time-lag bias. For (b) and (c) point size is proportional to the precision of the effect size (*i.e.* the inverse of its variance). For (a), the dashed red lines represent the 95 % CIs of the grand mean estimate for all effect sizes and the black dotted and dashed lines depict the 95 % and 99.8 % CIs for the dataset. The grey envelopes in (b) and (c) represent the 95 % CI of the linear regression. Source data are provided at \url{https://github.com/JustinCally/SexualSelection}.

\textbf{Figure 4: PRISMA diagram.} Flow of inclusion and exclusion of studies identified during the literature search, presented as a PRISMA diagram with number of published papers in brackets.
